How is experience transformed into memory?

### 2018-09-05 paper/main.tex
inputted andy's comments on main text

\begin{abstract}
How our experiences unfold over time define unique \textit{trajectories} through the relevant representational spaces.  Within this geometric framework, one can compare the shape of the trajectory formed by an experience to that defined by our later remembering of that experience.  We propose a framework for mapping naturalistic experiences onto geometric spaces that characterize how they unfold over time.  New insights emerge when we apply this approach to a naturalistic memory experiment which had participants view and recount a video.  We found that the shapes of the trajectories formed by participants' recountings were all highly similar to that of the original video, but participants differed in the level of detail they remembered.  We also identified a network of brain structures that are sensitive to the ``shapes'' of our ongoing experiences, and an overlapping network that is sensitive to how we will later remember those experiences.
\end{abstract}

@@

What does it mean to \textit{remember} something? In traditional episodic memory experiments \citep[e.g., list-learning or trial-based experiments;][]{Murd62a, Kaha96}, remembering is often cast as a discrete and binary operation: each studied item may be separated from the rest of one's experiences, and that item may be labeled as having been recalled versus forgotten. More nuanced studies might incorporate self-reported confidence measures as a proxy for memory strength, or ask participants to discriminate between ``recollecting'' the (contextual) details of an experience or having a general feeling of ``familiarity'' \citep{Yone02}. However, characterizing and evaluating memory in more realistic contexts (e.g., recounting a recent experience to a friend) is fundamentally different in at least three ways~\citep[also see][for a review]{KoriGold94}.  First, real world recall is continuous, rather than discrete.  Unlike in trial-based experiments, removing a (naturalistic) event from the context in which it occurs can substantially change its meaning.  Second, the specific words used to describe an experience have little bearing on whether the experience should be considered to have been ``remembered.''  Asking whether the rememberer has precisely reproduced a specific set of words to describe a given experience is nearly orthogonal to whether they were actually able to remember it.  In classic (e.g., list-learning) memory studies, counting the number or proportion of precise recalls is often a primary metric of assessing the quality of participants' memories.  Third, one might remember the \textit{gist} or essence of an experience but forget (or neglect to recount) particular details.  Capturing the gist of what happened is typically the main ``point'' of recounting a memory to a listener whereas, depending on the circumstances, accurate recall of any specific detail may be irrelevant.  There is no analog of the gist of an experience in most traditional memory experiments; rather we tend to assess participants' abilities to recover specific details (e.g., computing the proportion of specific stimuli they remember, which presentation positions the remembered stimuli came from, etc.).

@@

\cite{ChenEtal17} collected hand-annotated information about each of 1000 (manually identified) scenes spanning the roughly 45 minute video used in their experiment.  This information included: a brief narrative description of what was happening; whether the scene took place indoors vs. outdoors; names of any characters on the screen; names of any characters who were in focus in the camera shot; names of characters who were speaking; the location where the scene took place; the camera angle (close up, medium, long, etc.); whether or not background music was present; and other similar details (for a full list of annotated features see \textit{Methods}).  We took from these annotations the union of all unique words (excluding stop words, such as ``and,'' ``or,'' ``but,'' etc.) across all features and scenes as the ``vocabulary'' for the topic model.  We then concatenated the sets of words across all features contained in overlapping 50-scene sliding windows, and treated each 50-scene sequence as a single ``document'' for the purposes of fitting the topic model.  Next, we fit a topic model with (up to) $K = 100$ topics to this collection of documents.  We found that 28 unique topics (with non-zero weights) were sufficient to describe the time-varying content of the movie (see \textit{Methods}; Figs.~\ref{fig:schematic}, \topics).  Note that our approach is similar in some respects to Dynamic Topic Models~\citep{BleiLaff06}, in that we sought to characterize how the thematic content of the episode evolved over time.  However, whereas Dynamic Topic Models are designed to characterize how the properties of \textit{collections} of documents change over time, our sliding window approach allows us to examine the topic dynamics within a single document (or video).  Specifically, our approach yielded (via the topic proportions matrix) a single \textit{topic vector} for each timepoint of the episode (we set timepoints to match the acquisition times of the 1976 fMRI volumes collected as participants viewed the episode).

@@

\begin{figure}[tp]
\centering
\includegraphics[width=\textwidth]{figs/eventseg}
\caption{\small \textbf{Modelling naturalistic stimuli and recalls.} All panels: darker colors indicate greater values; range: [0, 1].  \textbf{A.} Topic vectors ($K = 100$) for each of the 1976 video timepoints.  \textbf{B.} Timepoint-by-timepoint correlation matrix of the topic vectors displayed in Panel A.  Event boundaries detected by the HMM are denoted in yellow (34 events detected).  \textbf{C.} Average topic vectors for each of the 34 video events. \textbf{D.} Topic vectors for each of 294 sentences spoken by an example participant while recalling the video.  \textbf{E.} Timepoint-by-timepoint correlation matrix of the topic vectors displayed in Panel D. Event boundaries detected by the HMM are denoted in yellow (27 events detected).  \textbf{F.} Average topic vectors for each of the 27 recalled events from the example participant.  \textbf{G.} Correlations between the topic vectors for every pair of video events (Panel C) and recalled events (from the example participant; Panel F).  For similar plots for all participants see Figure~\matchmats.  \textbf{H.} Average correlations between each pair of video events and recalled events (across all 17 participants).  To create the figure, each recalled event was assigned to the video event with the most correlated topic vector (yellow boxes in panels G and H).}
\label{fig:model}
\end{figure}

@@

Although the example participant's recall topic proportions matrix has some visual similarity to the video topic proportions matrix, the time-varying topic proportions for the example participant's recalls are not as sparse as for the video (e.g., compare Figs.~\ref{fig:model}A and D).  Similarly, although there do appear to be periods of stability in the recall topic dynamics (e.g., most topics are active or inactive over contiguous blocks of time), the overall timecourses are not as cleanly delineated as the video topics are.  To examine these patterns in detail, we computed the timepoint-by-timepoint correlation matrix for the example participant's recall topic proportions (Fig.~\ref{fig:model}E).  As in the video correlation matrix (Fig.~\ref{fig:model}B), the example partipant's recall correlation matrix has a strong block diagonal structure, indicating that their recalls are discretized into separated events.  As for the video correlation matrix, we can use an HMM, along with the aforementioned number-of-events optimization procedure (also see \textit{Methods}) to determine how many events are reflected in the participant's recalls and where specifically the event boundaries fall (outlined in yellow).  We carried out a similar analysis on all 17 participants' recall topic proportions matrices (Fig.~\corrmats).

@@

The above results indicate that both the structure of the original video and participants' recalls of the video exhibit event boundaries that can be identified automatically by characterizing the dynamic content using a shared topic model and segmenting the content into events using HMMs.  Next we asked whether some correspondence might be made between the specific content of the events the participants experienced in the video, and the events they later recalled.  One approach to linking the experienced (video) and recalled events is to label each recalled event as matching the video event with the most similar (i.e., most highly correlated) topic vector (Figs.~\ref{fig:model}G, \matchmats).  This yields a sequence of ``presented'' events from the original movie, and a sequence of (potentially differently ordered) ``recalled'' events for each participant.  Analogous to classic list-learning studies, we can then examine participants' recall sequences by asking which events they tended to recall first~\citep[probability of first recall; Fig.~\listlearning A;][]{WelcBurn24, PostPhil65, AtkiShif68}; how participants most often transition between recalls of the events as a function of the temporal distance between them~\citep[lag-conditional response probability; Fig.~\listlearning B;][]{Kaha96}; and which events they were likely to remember overall~\citep[serial position recall analyses; Fig.~\listlearning C;][]{Murd62a}.  In list-learning studies, this set of three analyses may be used to gain a nearly complete view into the sequences of recalls participants made~\citep[e.g., ][]{Kaha12}.  Extending these analyses to apply to naturalistic stimuli and recall~\citep{HeusEtal17b} highlights that, in naturalistic recall, these analyses provide a wholly incomplete picture: they leave out any attempt to quantify participants' abilities to capture the \textit{content} of what occurred in the video-- their only experimental instruction!

@@

\begin{figure}[tp]
\centering
\includegraphics[width=1\textwidth]{figs/trajectory}
\caption{\small \textbf{Trajectories through topic space capture the dynamic content of the video and recalls.}  All panels: the topic proportion matrices have been projected onto a shared two-dimensional space using UMAP.  \textbf{A.} The two-dimensional topic trajectory taken by the episode of \textit{Sherlock}.  Each dot indicates an event identified using the HMM (see \textit{Methods}); the dot colors denote the order of the events (early events are in red; later events are in blue), and the connecting lines indicate the transitions between successive events.  \textbf{B.} The average two-dimensional trajectory captured by participants' recall sequences, with the same format and coloring as the trajectory in Panel A.  To compute the event positions, we matched each recalled event with an event from the original video (see text), and then we averaged the positions of all events with the same label.  The arrows reflect the average transition direction through topic space taken by any participants whose trajectories crossed that part of topic space; blue denotes reliable agreement across participants via a Rayleigh test ($p < 0.05$, corrected).  \textbf{C.} The recall topic trajectories (gray) taken by each individual participant (P1--P17).  The video's trajectory is shown in black for reference.  (Same format and coloring as Panel A.)}
\label{fig:trajectory}
\end{figure}

Visual inspection of the video and recall topic trajectories reveals a striking pattern.  First, the topic trajectory of the video (which reflects its dynamic content; Fig.~\ref{fig:trajectory}A) is captured nearly perfectly by the averaged topic trajectories of participants' recalls (Fig.~\ref{fig:trajectory}B).  To assess the consistency of these recall trajectories across participants, we asked: given that a participant's recall trajectory had entered a particular location in topic space, could the position of their \textit{next} recalled event be predicted reliably?  For each location in topic space, we computed the set of line segments connecting successively recalled events (across all participants) that intersected that location (see \textit{Methods} for additional details).  We then computed (for each location) the distribution of angles formed by the lines defined by those line segments and a fixed reference line (the $x$-axis).  Rayleigh tests revealed the set of locations in topic space at which these across-participant distributions exhibited reliable peaks (blue arrows in Fig.~\ref{fig:trajectory}B).  We observed that the locations traversed by nearly the entire video trajectory exhibited such peaks.  In other words, participants exhibited similar trajectories that also matched the trajectory of the original video (Fig.~\ref{fig:trajectory}C).  This is especially notable when considering the fact that the number of events participants recalled (dots in Fig.~\ref{fig:trajectory}C) varied considerably across people, and that every participant used different words to describe what they had remembered happening in the video.  Differences in the numbers of remembered events appear in participants' trajectories as differences in the sampling resolution along the trajectory.  We note that this framework also provides a means of detangling classic ``proportion recalled'' measures (i.e., the proportion of video events referenced in participants' recalls) from participants' abilities to recapitulate the full gist of the original video (i.e., the similarity in the shape of the original video trajectory and that defined by each participant's recounting of the video).

Because our analysis framework projects the dynamic video content and participants' recalls onto a shared topic space, and because the dimensions of that space are known (i.e., each topic dimension is a set of weights over words in the vocabulary; Fig.~\topics), we can examine the topic trajectories to understand which specific content was remembered well (or poorly).  For each video event, we can ask: what was the average correlation (across participants) between the video event's topic vector and the closest matching recall event topic vectors from each participant? This yields a single correlation coefficient for each video event, describing how closely participants' recalls of the event tended to reliably capture its content (Fig.~\ref{fig:topics}A).  (We also examined how different comparisons between each video event's topic vector and the corresponding recall event topic vectors related to hand-annotated characterizations of memory performance; see \textit{Supporting information}).  Given this summary of which events were recalled reliably (or not), we next asked whether the better-remembered or worse-remembered events tended to reflect particular topics.  We computed a weighted average of the topic vectors for each video event, where the weights reflected how reliably each event was recalled.  To visualize the result, we created a ``wordle'' image~\citep{MuelEtal18} where words weighted more heavily by better-remembered topics appear in a larger font (Fig.~\ref{fig:topics}B, green box).  Events that reflected topics weighting heavily on characters like ``Sherlock'' and ``John'' (i.e., the main characters) and locations like ``221b Baker Street'' (i.e., a major recurring location and the address of the flat that Sherlock and John share) were best remembered.  An analogous analysis revealed which themes were poorly remembered.  Here in computing the weighted average over events' topic vectors we weighted each event in \textit{inverse} proportion to how well it was remembered (Fig.~\ref{fig:topics}B, red box).  This revealed that events with relatively minor characters such as ``Mike,'' ``Jeffrey,'' and ``Molly,'' as well as less-integral plot locations (e.g., ``hospital'' and ``office'') were least well-remembered.  This suggests that what is retained in memory are the major plot elements (i.e., the overall ``gist'' of what happened), whereas the more minor details are prone to pruning.

@@

The results thus far inform us about which aspects of the dynamic content in the episode participants watched were preserved or altered in participants' memories of the episode.  We next carried out a series of analyses aimed at understanding which brain structures might implement these processes.  In one analysis we sought to identify which brain structures were sensitive to the video's dynamic content, as characterized by its topic trajectory.  Specifically, we used a searchlight procedure to identify the extent to which each cluster of voxels exhibited a timecourse (as the participants watched the video) whose temporal correlation matrix matched the temporal correlation matrix of the original video's topic proportion matrix (Fig.~\ref{fig:model}B).  As shown in Figure~\ref{fig:brainz}A, the analysis revealed a network of regions including bilateral frontal cortex and cingulate cortex, suggesting that these regions may play a role in maintaining information relevant to the narrative structure of the video.  In a second analysis, we sought to identify which brain structures' responses (while viewing the video) reflected how each participant would later \textit{recall} the video.  We used an analogous searchlight procedure to identify clusters of voxels whose temporal correlation matrices reflected the temporal correlation matrix of the topic proportions for each individual's recalls (Figs.~\ref{fig:model}D, \corrmats).  As shown in Figure~\ref{fig:brainz}B, the analysis revealed a network of regions including the ventromedial prefrontal cortex (vmPFC), anterior cingulate cortex, and right medial temporal lobe (rMTL), suggesting that these regions may play a role in transforming each individual's experience into memory.

@@

One approach to identifying neural responses to naturalistic stimuli (including experiences) entails building a model of the stimulus and searching for brain regions whose responses are consistent with the model.  In prior work, a series of studies from Uri Hasson's group~\citep{LernEtal11, SimoEtal16, ChenEtal17, BaldEtal17, ZadbEtal17} have extended this approach with a clever twist.  Rather than building an explicit stimulus model, these studies instead search for brain responses (while experiencing the stimulus) that are reliably similar across individuals.  So called \textit{inter-subject correlation} (ISC) and \textit{inter-subject functional connectivity} (ISFC) analyses effectively treat other people's brain responses to the stimulus as a ``model'' of how its features change over time.  By contrast, in our present work we used topic models and HMMs to construct an explicit stimulus model (i.e., the topic trajectory of the movie).  When we searched for brain structures whose responses are consistent with the video's topic trajectory, we identified a network of structures that overlapped strongly with the ``long temporal receptive window'' network reported by the Hasson group~\citep[e.g., compare our Fig.~\ref{fig:brainz}A with the map of long temporal receptive window voxels in][]{LernEtal11}.  This provides support for the notion that part of the long temporal receptive window network may be maintaining an explicit model of the stimulus dynamics.  When we performed a similar analysis after swapping out the video's topic trajectory with the recall topic trajectories of each individual participant, this allowed us to identify brain regions whose responses (as the participants viewed the video) reflected how the video trajectory would be transformed in memory (as reflected by the recall topic trajectories).  The analysis revealed that the rMTL and vmPFC may play a role in this person-specific transformation from experience into memory.  The role of the MTL in episodic memory encoding has been well-reported~\citep[e.g., ][]{PallWagn02, DavaEtAl03, RangEtal04, Dava06}.  Prior work has also implicated the medial prefrontal cortex in representing ``schema'' knowledge~\citep[i.e., general knowledge about the format of an ongoing experience given prior similar experiences; ][]{KestEtal12, GilbMarl17}.  Integrating across our study and this prior work, one interpretation is that the person-specific transformations mediated (or represented) by the rMTL and vmPFC may reflect schema knowledge being leveraged, formed, or updated, incorporating ongoing experience into previously acquired knowledge.

@@

\subsection*{Experimental design and data collection}
Data were collected by \cite{ChenEtal17}.  In brief, participants ($n=17$) viewed the first 48 minutes of ``A Study in Pink'', the first episode of the BBC television series \textit{Sherlock}, while fMRI volumes were collected (TR = 1500~ms).  The stimulus was divided into a 23~min (946~TR) and a 25~min (1030~TR) segment to mitigate technical issues related to the scanner.  After finishing the clip, participants were instructed to \citep[quoting from][]{ChenEtal17} ``describe what they recalled of the movie in as much detail as they could, to try to recount events in the original order they were viewed in, and to speak for at least 10 minutes if possible but that longer was better. They were told that completeness and detail were more important than temporal order, and that if at any point they realized they had missed something, to return to it. Participants were then allowed to speak for as long as they wished, and verbally indicated when they were finished (e.g., `I’m done').''  For additional details about the experimental procedure and scanning parameters see \cite{ChenEtal17}.

After preprocessing the fMRI data and warping the images into a standard (3~mm$^3$ MNI) space, the voxel activations were $z$-scored (within voxel) and spatially smoothed using a 6~mm (full width at half maximum) Gaussian kernel.  The fMRI data were also cropped so that all video-viewing data were aligned across partipants.  This included a constant 3 TR (4.5~s) shift to account for the lag in the hemodynamic response.  \citep[All of these preprocessing steps followed][where additional details may be found.]{ChenEtal17}

>>>>

\begin{abstract}
How our experiences unfold over time define unique \textit{trajectories} through the relevant representational spaces.  Within this geometric framework, one can compare the shape of the trajectory formed by an experience to that defined by our later remembering of that experience.  We propose a framework for mapping naturalistic experiences onto geometric spaces that characterize how they unfold over time.  We apply this approach to a naturalistic memory experiment which had participants view and recount a video.  We found that the shapes of the trajectories formed by participants' recountings were all highly similar to that of the original video, but participants differed in the level of detail they remembered.  We also identified a network of brain structures that are sensitive to the ``shapes'' of our ongoing experiences, and an overlapping network that is sensitive to how we will later remember those experiences.
\end{abstract}

@@

What does it mean to \textit{remember} something? In traditional episodic memory experiments \citep[e.g., list-learning or trial-based experiments;][]{Murd62a, Kaha96}, remembering is often cast as a discrete and binary operation: each studied item may be separated from the rest of one's experiences, and that item may be labeled as having been recalled versus forgotten. More nuanced studies might incorporate self-reported confidence measures as a proxy for memory strength, or ask participants to discriminate between ``recollecting'' the (contextual) details of an experience or having a general feeling of ``familiarity'' \citep{Yone02}. However, characterizing and evaluating memory in more realistic contexts (e.g., recounting a recent experience to a friend) is fundamentally different in at least three ways~\citep[for review also see][]{KoriGold94, HukEtal18}.  First, real world recall is continuous, rather than discrete.  Unlike in trial-based experiments, removing a (naturalistic) event from the context in which it occurs can substantially change its meaning.  Second, the specific words used to describe an experience have little bearing on whether the experience should be considered to have been ``remembered.''  Asking whether the rememberer has precisely reproduced a specific set of words to describe a given experience is nearly orthogonal to whether they were actually able to remember it.  In classic (e.g., list-learning) memory studies, by contrast, counting the number or proportion of precise recalls is often a primary metric of assessing the quality of participants' memories.  Third, one might remember the \textit{gist} or essence of an experience but forget (or neglect to recount) particular details.  Capturing the gist of what happened is typically the main ``point'' of recounting a memory to a listener whereas, depending on the circumstances, accurate recall of any specific detail may be irrelevant.  There is no analog of the gist of an experience in most traditional memory experiments; rather we tend to assess participants' abilities to recover specific details (e.g., computing the proportion of specific stimuli they remember, which presentation positions the remembered stimuli came from, etc.).

@@

\cite{ChenEtal17} collected hand-annotated information about each of 1000 (manually identified) scenes spanning the roughly 45 minute video used in their experiment.  This information included: a brief narrative description of what was happening; whether the scene took place indoors vs. outdoors; names of any characters on the screen; names of any characters who were in focus in the camera shot; names of characters who were speaking; the location where the scene took place; the camera angle (close up, medium, long, etc.); whether or not background music was present; and other similar details (for a full list of annotated features see \textit{Methods}).  We took from these annotations the union of all unique words (excluding stop words, such as ``and,'' ``or,'' ``but,'' etc.) across all features and scenes as the ``vocabulary'' for the topic model.  We then concatenated the sets of words across all features contained in overlapping 50-scene sliding windows, and treated each 50-scene sequence as a single ``document'' for the purposes of fitting the topic model.  Next, we fit a topic model with (up to) $K = 100$ topics to this collection of documents.  We found that 27 unique topics (with non-zero weights) were sufficient to describe the time-varying content of the video (see \textit{Methods}; Figs.~\ref{fig:schematic}, \topics).  Note that our approach is similar in some respects to Dynamic Topic Models~\citep{BleiLaff06}, in that we sought to characterize how the thematic content of the episode evolved over time.  However, whereas Dynamic Topic Models are designed to characterize how the properties of \textit{collections} of documents change over time, our sliding window approach allows us to examine the topic dynamics within a single document (or video).  Specifically, our approach yielded (via the topic proportions matrix) a single \textit{topic vector} for each timepoint of the episode (we set timepoints to match the acquisition times of the 1976 fMRI volumes collected as participants viewed the episode).

@@

\begin{figure}[tp]
\centering
\includegraphics[width=\textwidth]{figs/eventseg}
\caption{\small \textbf{Modelling naturalistic stimuli and recalls.} All panels: darker colors indicate greater values; range: [0, 1].  \textbf{A.} Topic vectors ($K = 100$) for each of the 1976 video timepoints.  \textbf{B.} Timepoint-by-timepoint correlation matrix of the topic vectors displayed in Panel A.  Event boundaries detected by the HMM are denoted in yellow (34 events detected).  \textbf{C.} Average topic vectors for each of the 34 video events. \textbf{D.} Topic vectors for each of 294 sentences spoken by an example participant while recalling the video.  \textbf{E.} Timepoint-by-timepoint correlation matrix of the topic vectors displayed in Panel D. Event boundaries detected by the HMM are denoted in yellow (27 events detected).  \textbf{F.} Average topic vectors for each of the 27 recalled events from the example participant.  \textbf{G.} Correlations between the topic vectors for every pair of video events (Panel C) and recalled events (from the example participant; Panel F).  For similar plots for all participants see Figure~\matchmats.  \textbf{H.} Average correlations between each pair of video events and recalled events (across all 17 participants).  To create the figure, each recalled event was assigned to the video event with the most correlated topic vector (yellow boxes in panels G and H).  The heat maps in each panel were created using \texttt{Seaborn}~\citep{WaskEtal16}.}
\label{fig:model}
\end{figure}

@@

Although the example participant's recall topic proportions matrix has some visual similarity to the video topic proportions matrix, the time-varying topic proportions for the example participant's recalls are not as sparse as for the video (e.g., compare Figs.~\ref{fig:model}A and D).  Similarly, although there do appear to be periods of stability in the recall topic dynamics (e.g., most topics are active or inactive over contiguous blocks of time), the overall timecourses are not as cleanly delineated as the video topics are.  To examine these patterns in detail, we computed the timepoint-by-timepoint correlation matrix for the example participant's recall topic proportions (Fig.~\ref{fig:model}E).  As in the video correlation matrix (Fig.~\ref{fig:model}B), the example participant's recall correlation matrix has a strong block diagonal structure, indicating that their recalls are discretized into separated events.  As for the video correlation matrix, we can use an HMM, along with the aforementioned number-of-events optimization procedure (also see \textit{Methods}) to determine how many events are reflected in the participant's recalls and where specifically the event boundaries fall (outlined in yellow).  We carried out a similar analysis on all 17 participants' recall topic proportions matrices (Fig.~\corrmats).

@@

The above results indicate that both the structure of the original video and participants' recalls of the video exhibit event boundaries that can be identified automatically by characterizing the dynamic content using a shared topic model and segmenting the content into events using HMMs.  Next we asked whether some correspondence might be made between the specific content of the events the participants experienced in the video, and the events they later recalled.  One approach to linking the experienced (video) and recalled events is to label each recalled event as matching the video event with the most similar (i.e., most highly correlated) topic vector (Figs.~\ref{fig:model}G, \matchmats).  This yields a sequence of ``presented'' events from the original video, and a sequence of (potentially differently ordered) ``recalled'' events for each participant.  Analogous to classic list-learning studies, we can then examine participants' recall sequences by asking which events they tended to recall first~\citep[probability of first recall; Fig.~\listlearning A;][]{WelcBurn24, PostPhil65, AtkiShif68}; how participants most often transition between recalls of the events as a function of the temporal distance between them~\citep[lag-conditional response probability; Fig.~\listlearning B;][]{Kaha96}; and which events they were likely to remember overall~\citep[serial position recall analyses; Fig.~\listlearning C;][]{Murd62a}.  In list-learning studies, this set of three analyses may be used to gain a nearly complete view into the sequences of recalls participants made~\citep[e.g., ][]{Kaha12}.  Extending these analyses to apply to naturalistic stimuli and recall~\citep{HeusEtal17b} highlights that, in naturalistic recall, these analyses provide a wholly incomplete picture: they leave out any attempt to quantify participants' abilities to capture the \textit{content} of what occurred in the video-- their only experimental instruction!

@@

\begin{figure}[tp]
\centering
\includegraphics[width=1\textwidth]{figs/trajectory}
\caption{\small \textbf{Trajectories through topic space capture the dynamic content of the video and recalls.}  All panels: the topic proportion matrices have been projected onto a shared two-dimensional space using UMAP.  \textbf{A.} The two-dimensional topic trajectory taken by the episode of \textit{Sherlock}.  Each dot indicates an event identified using the HMM (see \textit{Methods}); the dot colors denote the order of the events (early events are in red; later events are in blue), and the connecting lines indicate the transitions between successive events.  \textbf{B.} The average two-dimensional trajectory captured by participants' recall sequences, with the same format and coloring as the trajectory in Panel A.  To compute the event positions, we matched each recalled event with an event from the original video (see \textit{Results}), and then we averaged the positions of all events with the same label.  The arrows reflect the average transition direction through topic space taken by any participants whose trajectories crossed that part of topic space; blue denotes reliable agreement across participants via a Rayleigh test ($p < 0.05$, corrected).  \textbf{C.} The recall topic trajectories (gray) taken by each individual participant (P1--P17).  The video's trajectory is shown in black for reference.  (Same format and coloring as Panel A.)}
\label{fig:trajectory}
\end{figure}

Visual inspection of the video and recall topic trajectories reveals a striking pattern.  First, the topic trajectory of the video (which reflects its dynamic content; Fig.~\ref{fig:trajectory}A) is captured nearly perfectly by the averaged topic trajectories of participants' recalls (Fig.~\ref{fig:trajectory}B).  To assess the consistency of these recall trajectories across participants, we asked: given that a participant's recall trajectory had entered a particular location in topic space, could the position of their \textit{next} recalled event be predicted reliably?  For each location in topic space, we computed the set of line segments connecting successively recalled events (across all participants) that intersected that location (see \textit{Methods} for additional details).  We then computed (for each location) the distribution of angles formed by the lines defined by those line segments and a fixed reference line (the $x$-axis).  Rayleigh tests revealed the set of locations in topic space at which these across-participant distributions exhibited reliable peaks (blue arrows in Fig.~\ref{fig:trajectory}B reflect significant peaks at $p < 0.05$, corrected).  We observed that the locations traversed by nearly the entire video trajectory exhibited such peaks.  In other words, participants exhibited similar trajectories that also matched the trajectory of the original video (Fig.~\ref{fig:trajectory}C).  This is especially notable when considering the fact that the number of events participants recalled (dots in Fig.~\ref{fig:trajectory}C) varied considerably across people, and that every participant used different words to describe what they had remembered happening in the video.  Differences in the numbers of remembered events appear in participants' trajectories as differences in the sampling resolution along the trajectory.  We note that this framework also provides a means of detangling classic ``proportion recalled'' measures (i.e., the proportion of video events referenced in participants' recalls) from participants' abilities to recapitulate the full gist of the original video (i.e., the similarity in the shape of the original video trajectory and that defined by each participant's recounting of the video).

Because our analysis framework projects the dynamic video content and participants' recalls onto a shared topic space, and because the dimensions of that space are known (i.e., each topic dimension is a set of weights over words in the vocabulary; Fig.~\topics), we can examine the topic trajectories to understand which specific content was remembered well (or poorly).  For each video event, we can ask: what was the average correlation (across participants) between the video event's topic vector and the closest matching recall event topic vectors from each participant? This yields a single correlation coefficient for each video event, describing how closely participants' recalls of the event tended to reliably capture its content (Fig.~\ref{fig:topics}A).  (We also examined how different comparisons between each video event's topic vector and the corresponding recall event topic vectors related to hand-annotated characterizations of memory performance; see \textit{Supporting Information}).  Given this summary of which events were recalled reliably (or not), we next asked whether the better-remembered or worse-remembered events tended to reflect particular topics.  We computed a weighted average of the topic vectors for each video event, where the weights reflected how reliably each event was recalled.  To visualize the result, we created a ``wordle'' image~\citep{MuelEtal18} where words weighted more heavily by better-remembered topics appear in a larger font (Fig.~\ref{fig:topics}B, green box).  Events that reflected topics weighting heavily on characters like ``Sherlock'' and ``John'' (i.e., the main characters) and locations like ``221b Baker Street'' (i.e., a major recurring location and the address of the flat that Sherlock and John share) were best remembered.  An analogous analysis revealed which themes were poorly remembered.  Here in computing the weighted average over events' topic vectors we weighted each event in \textit{inverse} proportion to how well it was remembered (Fig.~\ref{fig:topics}B, red box).  This revealed that events with relatively minor characters such as ``Mike,'' ``Jeffrey,'' and ``Molly,'' as well as less-integral plot locations (e.g., ``hospital'' and ``office'') were least well-remembered.  This suggests that what is retained in memory are the major plot elements (i.e., the overall ``gist'' of what happened), whereas the more minor details are prone to pruning.

@@

The results thus far inform us about which aspects of the dynamic content in the episode participants watched were preserved or altered in participants' memories of the episode.  We next carried out a series of analyses aimed at understanding which brain structures might implement these processes.  In one analysis we sought to identify which brain structures were sensitive to the video's dynamic content, as characterized by its topic trajectory.  Specifically, we used a searchlight procedure to identify the extent to which each cluster of voxels exhibited a timecourse (as the participants watched the video) whose temporal correlation matrix matched the temporal correlation matrix of the original video's topic proportion matrix (Fig.~\ref{fig:model}B).  As shown in Figure~\ref{fig:brainz}A, the analysis revealed a network of regions including bilateral frontal cortex and cingulate cortex, suggesting that these regions may play a role in maintaining information relevant to the narrative structure of the video.  In a second analysis, we sought to identify which brain structures' responses (while viewing the video) reflected how each participant would later \textit{recall} the video.  We used an analogous searchlight procedure to identify clusters of voxels whose temporal correlation matrices reflected the temporal correlation matrix of the topic proportions for each individual's recalls (Figs.~\ref{fig:model}D, \corrmats).  As shown in Figure~\ref{fig:brainz}B, the analysis revealed a network of regions including the ventromedial prefrontal cortex (vmPFC), anterior cingulate cortex, and right medial temporal lobe (rMTL), suggesting that these regions may play a role in transforming each individual's experience into memory.  In identifying regions whose responses to ongoing experiences reflect how those experiences will be remembered later, this latter analysis extends classic \textit{subsequent memory analyses}~\citep[e.g.,][]{PallWagn02} to domain of naturalistic stimuli.

@@

Our work casts remembering as reproducing (behaviorally and neurally) the topic trajectory, or ``gist,'' of the original experience.  This view draws inspiration from prior work aimed at elucidating the neural and behavioral underpinnings of how we process dynamic naturalistic experiences and remember them later.  One approach to identifying neural responses to naturalistic stimuli (including experiences) entails building a model of the stimulus and searching for brain regions whose responses are consistent with the model.  In prior work, a series of studies from Uri Hasson's group~\citep{LernEtal11, SimoEtal16, ChenEtal17, BaldEtal17, ZadbEtal17} have extended this approach with a clever twist.  Rather than building an explicit stimulus model, these studies instead search for brain responses (while experiencing the stimulus) that are reliably similar across individuals.  So called \textit{inter-subject correlation} (ISC) and \textit{inter-subject functional connectivity} (ISFC) analyses effectively treat other people's brain responses to the stimulus as a ``model'' of how its features change over time.  By contrast, in our present work we used topic models and HMMs to construct an explicit stimulus model (i.e., the topic trajectory of the video).  When we searched for brain structures whose responses are consistent with the video's topic trajectory, we identified a network of structures that overlapped strongly with the ``long temporal receptive window'' network reported by the Hasson group~\citep[e.g., compare our Fig.~\ref{fig:brainz}A with the map of long temporal receptive window voxels in][]{LernEtal11}.  This provides support for the notion that part of the long temporal receptive window network may be maintaining an explicit model of the stimulus dynamics.  When we performed a similar analysis after swapping out the video's topic trajectory with the recall topic trajectories of each individual participant, this allowed us to identify brain regions whose responses (as the participants viewed the video) reflected how the video trajectory would be transformed in memory (as reflected by the recall topic trajectories).  The analysis revealed that the rMTL and vmPFC may play a role in this person-specific transformation from experience into memory.  The role of the MTL in episodic memory encoding has been well-reported~\citep[e.g., ][]{PallWagn02, DavaEtAl03, RangEtal04, Dava06}.  Prior work has also implicated the medial prefrontal cortex in representing ``schema'' knowledge~\citep[i.e., general knowledge about the format of an ongoing experience given prior similar experiences; ][]{KestEtal12, SchlPres15, GilbMarl17, SpalEtal18}.  Integrating across our study and this prior work, one interpretation is that the person-specific transformations mediated (or represented) by the rMTL and vmPFC may reflect schema knowledge being leveraged, formed, or updated, incorporating ongoing experience into previously acquired knowledge.

@@

\subsection*{Experimental design and data collection}
Data were collected by \cite{ChenEtal17}.  In brief, participants ($n=17$) viewed the first 48 minutes of ``A Study in Pink'', the first episode of the BBC television series \textit{Sherlock}, while fMRI volumes were collected (TR = 1500~ms).  The stimulus was divided into a 23~min (946~TR) and a 25~min (1030~TR) segment to mitigate technical issues related to the scanner.  After finishing the clip, participants were instructed to \citep[quoting from][]{ChenEtal17} ``describe what they recalled of the [episode] in as much detail as they could, to try to recount events in the original order they were viewed in, and to speak for at least 10 minutes if possible but that longer was better. They were told that completeness and detail were more important than temporal order, and that if at any point they realized they had missed something, to return to it. Participants were then allowed to speak for as long as they wished, and verbally indicated when they were finished (e.g., `I’m done').''  For additional details about the experimental procedure and scanning parameters see \cite{ChenEtal17}.

After preprocessing the fMRI data and warping the images into a standard (3~mm$^3$ MNI) space, the voxel activations were $z$-scored (within voxel) and spatially smoothed using a 6~mm (full width at half maximum) Gaussian kernel.  The fMRI data were also cropped so that all video-viewing data were aligned across participants.  This included a constant 3 TR (4.5~s) shift to account for the lag in the hemodynamic response.  \citep[All of these preprocessing steps followed][where additional details may be found.]{ChenEtal17}
